---
title: "Stochastic DTMC"
author: "Vidya Mallick"
date: "01/12/2021"
output: word_document
---

DTMC
a. Take input as the TPM for irreducible markov chain
b. Find Steady State Behavior by solving equations.
c. Simulate this for multiple runs
d. Find behavior from the simulation perspective 
e. Identification of recurrent / transient states  and classes 


WEATHER PROBLEM
 
Problem Statement: It is given that, if today is rainy day then the probability that tomorrow will be a rainy day is 0.5 and the chances to be nice and snowy weather are respectively 0.25 and 0.25. The chances for rainy, nice and snowy weather if today is a nice day are 0.5,0 and 0.5. If today is a snowy day the chances for rainy, nice and snowy weather are 0.25,0.25 and 0.5 respectively. 


```{r}
#install.packages("markovchain")
library(markovchain)
```
## To find the TPM:

```{r}
WeatherStatespace=c("Rainy","Sunny")
WeatherTPM=matrix(c(0.7,0.3,0.2,0.8),nrow=2,byrow=TRUE,dimnames=list(WeatherStatespace,WeatherStatespace))
WeatherTPM
```
The above matrix shows the Transition Probability matrix for Weather Problem.

## Creating Markovchain object using markovchain function

```{r}
WeatherMC=new("markovchain",states=WeatherStatespace,byrow=T,transitionMatrix=WeatherTPM,name="Weather problem")
WeatherMC
```
## Check if its Irreducible:

```{r}
is.irreducible(WeatherMC)
```
The Markov Chain - "WeatherMC" is irreducible.

## Plotting the transition diagram

```{r}
#install.packages("diagram")
library(diagram)
```

```{r}
## Loading required package: shape
plotmat(WeatherTPM,pos=c(1,1),lwd=1,box.lwd = 1,cex.tex=0.5,box.size = 0.1,box.type = "circle",box.prop = 0.4,box.col = "light yellow",arr.length = 0.1,arr.width = 0.1,self.cex = 0.4,self.shifty = -0.01,self.shiftx = 0.13,main = "Transition Diagram")
```

The above diagram shows the transition between the various states i.e between Rainy and Sunny.

## Checking Accessibility

```{r}
is.accessible(WeatherMC)
```
## Communicative States

```{r}
communicatingClasses(WeatherMC)
```

## Finding Higher Transition probabilities

## 2-step and 5-step TPMs

2-Step

```{r}
WeatherMC^2
```

Inference: The above shows that if today is a rainy day then in a two step process the probability for tomorrow to be a rainy day is explained as follows:
Probability = P(Rainy today & tomorrow)+ P(Rainy today x Sunny Tomorrow) = (0.7 x 0.7) + (0.3 x 0.2 = 0.55
Similarly, the same can be applied for 5 step, the output is shown below.

5-Step

```{r}
WeatherMC^5
```
Inference: 

## Steady state probabilities:

```{r}
steadyStates(WeatherMC)
```

From the above table we can see that with a large number of transitions the probability of the weather at steady state will be: Chances of being a rainy day is 40% Chances of being a nice day is 20% Chances of being a Snowy day is 40%

## Probability that Wednesday to be Sunny if Monday is a rainy day

```{r}
WeatherMC^2
```

From the above table the prabability that Wednesday is a Snowy if Monday is Rainy day is 0.45. This is because there are 2 transitions . From monday to tuesday and from tuesday to wednesday. In the so formed 2 square transition matrix we choose 1st row 3rd column to find probability of snowy day if it is rainy.

## Probability of weather to be either "Rainy" or "Sunny" after 30 days based on initial probability.

```{r}
initial=c(0.5,0.5)
prob_30 = initial*WeatherMC^30
prob_30
```
The above matrix shows how the weather would behave after 30 days. We can see that, there is a 40% chance for the weather to be "Rainy" and a 60% chance for the weather to be "Sunny".

## Probability that it is rainy after 12 days when the initial probability is given

```{r}
initial=c(0.5,0.5)
prob_12 = initial*WeatherMC^12
prob_12[1]
```
There is 40% chance that it will be raining after 12 days if the initial probability is given.

## Identification of recurrent / transient states  and classes

#To find the recurrent states

```{r}
recurrentStates(WeatherMC)
```


#To find the transient States

```{r}
transientStates(WeatherMC)
```


#To find the recurrent Classes

```{r}
recurrentClasses(WeatherMC)
```

#To find the transient Classes

```{r}
transientClasses(WeatherMC)
```

#summary of DTMC

```{r}
summary(WeatherMC)
```

#simulation

```{r}
#install.packages("tibble")
library(tibble)
```


```{r}
weathersOfDays <- rmarkovchain(n = 365, object = WeatherMC,initial)
weathersOfDays 
weathersOfDays[1:30]
enframe(weathersOfDays)
```


## Conclusion:

We concluded that along with finding transition probability matrix one can also obtain the steady states policy and also find dependent probability . We also found 2 and 5 step transition probability matrix. This shows r can help in formulating markov chain problems


## STOCK MARKET

Problem Statement: An open stock market is exhibiting a bull market, bear market, or stagnant market trend during a given week. A bull week is followed by another bull week 90% of the time, a bear week 7.5% of the time, and a stagnant week the other 2.5% of the time. A bear week is followed by another bear week 80% of the time, a bull week 15% of the time and a stagnant week 5% of the time. There is a 25% chance that a stagnant week is followed by a bull week and 50% for followed by a bear week.

## AIM: To understand the changes in the Stock Market using Markov Chain and the theory of probabilities in Stochastic Processes. 

## PROCEDURE:

## TPM and Markov Chain

```{r}
MarketStatespace=c("Bull","Bear","Stagnant")
MarketTPM=matrix(c(0.9,0.075,0.025,0.15,0.8,0.05,0.25,0.5,0.25),nrow=3,byrow=TRUE,dimnames=list(MarketStatespace,MarketStatespace))
MarketTPM
```

```{r}
MarketMC=new("markovchain",states=MarketStatespace,byrow=T,transitionMatrix=MarketTPM,name="Market problem")
MarketMC

```

## Communicating Classes

```{r}
communicatingClasses(MarketMC)
```
## Irreducibility

```{r}
is.irreducible(MarketMC)
```

## probability that the market is stagnant after 7 weeks

```{r}
initial=c(0.2,0.5,0.3)
prob_7 = initial*MarketMC^7
prob_7[3]

```
There is a 4.5% that the market is stagnant after 7 weeks.

# State Probabilities at nth step when initial probability is given

If the initial probability distribution is given by (0.2,0.5,0.3) then find the probability that the market is exhibiting a bear market after 4 weeks given that initially, the market is stagnant.

```{r}
x0 = 0.2 #bull 
x1 = 0.5 #bear
x2 = 0.3 #stagnant

init_4 = conditionalDistribution(MarketMC^4,"Stagnant")[2]
init_4

```
Initial state of the market being at Stagnant, the probability that after 4 weeks the market is Bear is 0.4694375. Therefore there is a 46.9% chance.

# Suppose the market exhibits a bull market in the initial week find the joint probability that the market exhibits a bull market after 6 weeks and a stagnant market after 12 weeks.

```{r}
Pr=initial[1]*conditionalDistribution(MarketMC^6,"Bull")[1]*
conditionalDistribution(MarketMC^12,"Bull")[3]
Pr
```
Given Initial market state as “Bull”, the probability that after 6 weeks going to Bull and then after 12 weeks reaching Stagnant market is 0.005825353

## Identification of recurrent / transient states  and classes

#To find the recurrent states

```{r}
recurrentStates(MarketMC)
```


#To find the transient States

```{r}
transientStates(MarketMC)
```


#To find the recurrent Classes

```{r}
recurrentClasses(MarketMC)
```

#To find the transient Classes

```{r}
transientClasses(MarketMC)
```

#summary of DTMC

```{r}
summary(MarketMC)
```

#simulation

```{r}
Market_w <- rmarkovchain(n = 365, object = MarketMC,initial)
Market_w 
Market_w[1:30]
enframe(Market_w)
```


## Conclusion:

From the above procedures we can find the changes in the Stock Market from the three different states, “Bull” , “Bear” and “Stagnant”. And we can predict the probability of a market state given the initial probability.


